A modal logical multiverse for commutative algebra and combinatorics

### 2022-11-05 slides-paris2022.tex
Slides almost finished

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\BB}{\mathbb{B}}
\newcommand{\pp}{\mathbbm{p}}
\newcommand{\MM}{\mathbb{M}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\FF}{\mathbb{F}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LocRing}{\mathrm{LocRing}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ul{\ensuremath{\AA}}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates

@@

\newcommand{\badbox}[1]{\colorbox{red!30}{#1}}

@@

@@

>>>>

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\BB}{\mathbb{B}}
\newcommand{\pp}{\mathbbm{p}}
\newcommand{\MM}{\mathbb{M}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\FF}{\mathbb{F}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\ccc}{\mathfrak{c}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LocRing}{\mathrm{LocRing}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ul{\ensuremath{\AA}}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates

@@

\newcommand{\badbox}[1]{\colorbox{red!30}{#1}}
\newcommand{\infobox}[1]{\colorbox{yellow!70}{\color{black}#1}}

@@

\section{The set-theoretic multiverse}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.59cm}\includegraphics[width=\paperwidth]{staircase}\end{minipage}}
\begin{frame}{A brief timeline}
  \begin{itemize}
    \item[18xx] Cantor studies the \hil{continuum hypothesis}, the claim
    that~$2^{\aleph_0} = \aleph_1$.
    \item[19xx] Zermelo and Fraenkel offer the axioms of~\textsc{zfc}.
    \item[1938] Gödel proves: If~\textsc{zfc} is consistent, then so
    is~\textsc{zfc}+\textsc{ch}.
    \item[1963] Cohen proves: If~\textsc{zfc} is consistent, then so
    is~\textsc{zfc}+$\neg$\textsc{ch}.
    \item[since 1963 and ongoing] Set theorists pursue additional axioms to
    settle~\textsc{ch} (one way or another).  % XXX persue?
    \item[201x] Hamkins offers his paper on the \hil{multiverse position} in
    the philosophy of set theory, arguing that the program of pursing
    additional axioms (while successful in many ways) is doomed to fail in
    settling~\textsc{ch}.
  \end{itemize}
\end{frame}
\begin{frame}{The set-theoretic multiverse}
  \begin{tikzpicture}[remember picture,overlay]
    \node[xshift=-1.5cm,yshift=-3.5cm] at (current page.north east)
    {\includegraphics[width=2cm]{branching}};
  \end{tikzpicture}
  \vspace*{-1.5em}

  \begin{block}{}
    \justifying
    \textbf{Def.} A \hil{model of set theory} is a (perhaps class-sized) structure $(M,{\in})$
    satisfying axioms such as those of~\textsc{zfc}.
  \end{block}
  \vspace*{-0.6em}
  {\small\emph{Examples.}\vspace*{-0.9em}
  \begin{itemize}
    \item $V$, the class of all sets \vspace*{-0.6em}
    \item $L$, Gödel's constructible universe \vspace*{-0.6em}
    \item $V[G]$, a forcing extension containing a generic filter~$G$ of \\ some
    poset of forcing conditions \vspace*{-0.6em}
    \item Henkin/term models from consistency of (extensions of)~\textsc{zfc}
  \end{itemize}}
  \pause

  %{\centering\emph{Embracing all models of set theory:}\par}
  \begin{block}{}
    \justifying
    \textbf{Def.} $\possible\varphi$ iff~$\varphi$ holds in \hil{some extension} of
    the current universe. \\
    \phantom{\textbf{Def.}} $\necessary\varphi$ iff~$\varphi$ holds in \hil{all extensions}
    of the current universe.
  \end{block}
  \vspace*{-0.6em}
  \begin{itemize}
    \item $\necessary(\possible\textsc{CH} \wedge \possible\neg\text{CH})$,
    the continuum hypothesis is a \hil{switch}
    \item \ \\[-1.2em]\mbox{$\necessary\possible\necessary(\text{$X$ is countable})$,
    existence of an enumeration is a \hil{button}}
  \end{itemize}
  \pause
  \medskip

  \textbf{NB.} In the multiverse of extensions of a given field~$K$,

  \vspace*{-0.5em}
  \begin{itemize}
    \item ``there is a square root of~$-1$'' is a button and
    \item ``every polynomial splits into linear factors'' is a switch.
  \end{itemize}
\end{frame}}

\section{Parametrized mathematics}

\renewcommand{\insertframeextra}{a}
\begin{frame}{Parametrized mathematics}
  \medskip
  {\centering\includegraphics[width=0.7\textwidth]{eigenvector}\par}
  \small

  \begin{changemargin}{-0.5em}{-1.5em}
  \begin{enumerate}
    \item[\infobox{local}]
    \ \\[-1.2em]\mbox{``Every real symmetric matrix
    does \only<4>{\hil{not not} }have an eigenvector.''
    \good{\cmark}}
    \medskip

    \item[\infobox{global}] ``For every continuous family of symmetric
    matrices, \\\only<4>{\hil{on a dense open} }eigenvectors can locally
    be picked continuously.''
    \only<1-2>{\includegraphics[height=0.7em]{question-mark}}%
    \only<3>{\bad{\xmark}}%
    \only<4>{\good{\cmark}}

    \smallskip
    \pause
    \justifying
    \scriptsize
    ``Let~$X$ be a topological space and let~$A : X \to
    M^\text{sym}_n(\RR)$ be a continuous map to the space of symmetric~$(n \times
    n)$-matrices. Then there is an open covering~$\bigcup_{i \in I} U_i$
    \only<1-3>{of~$X$}\only<4>{\hil{of a dense open subset}~$U \subseteq X$} such
    that or all indices~$i \in I$, there is a continuous map~$v : U_i \to \RR^n$
    such that for all~$x \in U_i$, the vector~$v(x)$ is an eigenvector
    of~$A(x)$.''
  \end{enumerate}
  \end{changemargin}
\end{frame}

\addtocounter{framenumber}{-1}
\renewcommand{\insertframeextra}{b}
\begin{frame}{Parametrized mathematics}
  \bigskip
  \bigskip
  {\centering
    \includegraphics[height=9em]{damian-gvirtz-moduli-1}
    \quad
    \includegraphics[height=9em]{damian-gvirtz-moduli-2}
    \par}
  \small
  \bigskip

  \begin{changemargin}{-0.5em}{-1.5em}
  \begin{enumerate}
    \item[\infobox{local}]
    \ \\[-1.2em]``Let~$M$ be a finitely generated module over a
    field~$k$. \\ Then~$M$ is \only<4>{\hil{not not} }finite free.''
    \good{\cmark}
    \medskip

    \item[\infobox{global}] ``Let~$M$ be a finitely generated module over a
    ring~$A$. \\ Then~$M^\sim$ is finite locally
    free\only<4>{ \hil{on a dense open}}.''
    \only<1-2>{\includegraphics[height=0.7em]{question-mark}}%
    \only<3>{\bad{\xmark}}%
    \only<4>{\good{\cmark}}

    \smallskip
    \pause
    \justifying
    \scriptsize
    ``Let~$M$ be a finitely generated module over an
    arbitrary commutative ring~$A$. \only<1-3>{Then there is a partition~$1 = f_1 + \cdots +
    f_n \in A$ of unity such that, for each index~$i$, the localized
    module~$M[f_i^{-1}]$ is finite free over~$A[f_i^{-1}]$.}\only<4>{\hil{If~$f = 0$
    is the only element of~$A$} such that~$M[f^{-1}]$ is finite free
    over~$A[f^{-1}]$, \hil{then~$1 = 0$ in~$A$}.}''
  \end{enumerate}
  \end{changemargin}
\end{frame}

\addtocounter{framenumber}{-1}
\renewcommand{\insertframeextra}{c}
{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-1em}\includegraphics[width=\paperwidth]{train-tracks-cropped}\end{minipage}}
\begin{frame}{Parametrized mathematics}
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \small

  \begin{changemargin}{-0.5em}{-1.5em}
  \begin{enumerate}
    \item[\infobox{local}]
    \ \\[-1.2em]``Let $R$ be a ring.
    Let $n \geq 0$ be an integer.
    We have
    $$
    H^q(\mathbf{P}^n, \mathcal{O}_{\mathbf{P}^n_R}(d)) = \textit{(omitted)}
    $$
    as $R$-modules.''
    \good{\cmark}
    \medskip

    \item[\infobox{global}] ``Let $S$ be a scheme. Let $n \geq 1$.
    Let $\mathcal{E}$ be a finite locally
    free $\mathcal{O}_S$-module of constant rank $n + 1$.
    For the structure morphism
    $
    \pi : \mathbf{P}(\mathcal{E}) \longrightarrow S,
    $
    we have
    $$
    R^q\pi_*(\mathcal{O}_{\mathbf{P}(\mathcal{E})}(d)) =
    \textit{(omitted)}
    $$
    as sheaves of~$\mathcal{O}_S$-modules.'' \good{\cmark}
  \end{enumerate}
  \end{changemargin}
\end{frame}}

\renewcommand{\insertframeextra}{}
{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
\begin{frame}{Sites and toposes}
  For every \hil{site}~$\C$, the
  \hil{sheaves} on~$\C$ form a \hil{topos}. \emph{Notable sites:}
  \begin{enumerate}
    \item The \hil{site of opens} of a topological space
    \item The \hil{Zariski site} of a ring
    \item The \hil{site of finite approximations} of surjections~$\NN
    \twoheadrightarrow X$
    \item The \hil{classifying site} of a geometric theory
  \end{enumerate}
  \pause

  Every topos supports an \hil{internal language},
  and this language is sound with respect to \hil{intuitionistic
  logic}.
  \pause

  \begin{block}{}
    \justifying
    \textbf{Def.} A statement~$\varphi$ holds \ldots
    \begin{itemize}
      \item\small \hil{everywhere} ($\necessary\varphi$) iff
      it holds in every (Grothendieck) topos (over the current base topos).
      \item \hil{somewhere} ($\possible\varphi$) iff
      it holds in some positive topos.
      \item \hil{proximally} ($\xpossible\varphi$) iff
      it holds in some positive ouvert topos.
    \end{itemize}
  \end{block}
\end{frame}
\begin{frame}{The multiverse of parametrized mathematics}
  \begin{block}{}
    \justifying
    \textbf{Def.} A statement~$\varphi$ holds \ldots
    \begin{itemize}
      \item\small \hil{everywhere} ($\necessary\varphi$) iff
      it holds in every (Grothendieck) topos (over the current base topos).
      \item \hil{somewhere} ($\possible\varphi$) iff
      it holds in some positive topos.
      \item \hil{proximally} ($\xpossible\varphi$) iff
      it holds in some positive ouvert topos.
    \end{itemize}
  \end{block}

  \begin{changemargin}{-1.2em}{-1.2em}
    \small
    \begin{enumerate}
      %\item If a function~$\NN \to \NN$ does \emph{everywhere} \emph{not~not} have
      %a zero, then it actually has a zero.
      \item \emph{Somewhere,} the law of
      excluded middle holds. \emph{In fact, we even have:} \\
      It is \emph{everywhere} the case that the law of
      excluded middle holds \emph{somewhere}.
      \smallskip
      \item Assuming Zorn's lemma, it is
      everywhere the case that it the axiom of choice holds
      somewhere. [Barr]
      \smallskip
      \item If a geometric implication holds \emph{somewhere}, then it holds already here.
      \smallskip
      \item If a first-order statement holds \emph{proximally}, then it holds already here.
      \smallskip
      \item For every set~$X$, \emph{proximally} there is a
      surjection~$\NN \twoheadrightarrow X$.
      \smallskip
      \item Every ring \emph{proximally} has a maximal ideal.
    \end{enumerate}
  \end{changemargin}
\end{frame}}

@@

Menagerie of properties of the multiverse
